---
title: "Getting started with Enroll-HD data"
subtitle: "A tutorial for statisticians, in R and the tidyverse<br><span class='disclaimer'>Built only from publicly available Enroll-HD documentation. No real participant data are used or shown; every table and figure is illustrative.</span>"
author: "Abhishek"
date: today
title-slide-attributes:
  data-notes: "Disclaimer: public sources only, no participant data."
format:
  revealjs:
    theme: [simple, custom.scss]
    slide-number: true
    code-line-numbers: false
    code-overflow: wrap
    footer: "Enroll-HD for statisticians"
    width: 1280
    height: 720
    fig-width: 8
    fig-height: 4
    fig-dpi: 150
    toc: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(gt)
library(scales)
source(here::here("R", "make_toy_pds.R"))
theme_set(theme_minimal(base_size = 14))
pds <- make_toy_pds()
```

## Disclaimer {.disclaimer-slide}

::: callout-important
## Public information only, no real data
- This deck is built **only from publicly available material**: the Enroll-HD documentation set published for researchers (protocol, data dictionary, annotated case report forms, data-collection guidelines, dataset overview and interpretation guides) and the *Analyzing Data* articles at enroll-hd.org.
- **No Enroll-HD participant data are used, shown or summarised.** Every table printed by the code is a small **illustrative** dataset generated in `R/make_toy_pds.R`; every histogram or curve is either synthetic or an approximate redrawing of a figure already published on enroll-hd.org, and is labelled as such.
- Numbers quoted in the text (cohort size, distributions, retention, cut points) are taken from those public documents and articles and are cited for orientation, not for analysis.
- Nothing here replaces the official documentation or the data use agreement that governs access to a periodic dataset.
:::

## What this deck is for

::: {.incremental}
- You are a statistician who has just been given (or is about to request) an Enroll-HD periodic dataset.
- In the next hour you will learn **what the disease and the study are**, **how the files fit together**, **the conventions that bite**, and **how to start an analysis in tidy R**.
- Every code chunk runs on tiny illustrative tables built in `R/make_toy_pds.R`. No participant data are used or shown.
- Sources: the Enroll-HD documentation set and the *Analyzing Data* articles at enroll-hd.org. Links are on the last slide.
:::

## Agenda

1. Huntington's disease in five minutes
2. The Enroll-HD platform and its periodic datasets
3. Anatomy of a PDS: files, keys, visits
4. Loading and merging with the tidyverse
5. De-identification conventions and missing values
6. Disease definitions and derived scores
7. Special characteristics of HD data
8. Age, CAG and time in models
9. Coded therapies and comorbidities
10. Data quality and a pre-analysis checklist

# 1. Huntington's disease in five minutes

## The gene and the repeat

::: columns
::: {.column width="55%"}
- Autosomal dominant, caused by a CAG repeat expansion in *HTT*.
- The **larger allele** (`caghigh` in the data) is what matters:

| CAG repeats | Meaning |
|---|---|
| 10 to 26 | normal |
| 27 to 35 | intermediate allele |
| 36 to 39 | reduced penetrance |
| 40 and above | full penetrance |
| above 60 | usually juvenile onset |

- Modal expanded CAG in Enroll-HD is **42**; most carriers sit between 40 and 44. Modal non-expanded allele is **17**.
:::
::: {.column width="45%"}
- Longer repeats mean earlier onset, but with wide variation at any given length.
- Every participant is genotyped at a central research lab. The result is never returned to participants or sites; it is for research only.
- "HDGEC" = HD gene expansion carrier (36+). Controls are family members without the expansion.
:::
:::

## One disease, several clocks

Onset is not one event. Different domains start at different times, and the data hold several onset variables.

```{r}
#| label: natural-history
#| echo: false
#| fig-height: 4.2
# Approximate proportion-of-maximum trajectories for a CAG-42 cohort,
# redrawn from the Enroll-HD "Special characteristics of HD data" article.
natural_history <- tribble(
  ~age, ~Chorea, ~`SDMT deficit`, ~`Total motor score`, ~`TFC retained`,
   20,   0.00,    0.00,             0.00,                 1.00,
   35,   0.03,    0.02,             0.01,                 0.99,
   45,   0.14,    0.11,             0.04,                 0.96,
   50,   0.28,    0.21,             0.08,                 0.92,
   60,   0.60,    0.45,             0.25,                 0.75,
   70,   0.88,    0.70,             0.57,                 0.43,
   80,   1.00,    1.00,             1.00,                 0.00
) |>
  pivot_longer(-age, names_to = "measure", values_to = "proportion")

milestones <- tribble(
  ~age, ~label,
   34,  "first change",
   44,  "TFC < 13",
   50,  "motor diagnosis\n(DCL = 4)",
   59,  "stage 3",
   69.5, "stage 4+"
)

ggplot(natural_history, aes(age, proportion, colour = measure)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 1.1) +
  geom_vline(data = milestones, aes(xintercept = age), linetype = "dotted") +
  geom_text(data = milestones, aes(x = age, y = 1.08, label = label),
            size = 3, inherit.aes = FALSE) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1.15),
                     breaks = seq(0, 1, .25)) +
  labs(x = "Age (years), CAG = 42", y = "Proportion of maximum",
       colour = NULL,
       caption = "Approximate, redrawn from enroll-hd.org (PDS4). Not for analysis.") +
  theme(legend.position = "bottom")
```

## Two definitions of "manifest"

::: columns
::: {.column width="50%"}
**Motor diagnosis** (the literature's "manifest")

- Diagnostic Confidence Level `diagconf` = 4: motor signs are unequivocal signs of HD (99% confidence).
- Derived from the UHDRS motor exam at each visit.
- The **first visit** at which `diagconf` becomes 4 is the motor-onset date.
:::
::: {.column width="50%"}
**Participant category** `hdcat`

| code | category |
|---|---|
| 2 | premanifest / premotor-manifest |
| 3 | manifest / motor-manifest |
| 4 | genotype negative |
| 5 | family control |

- Clinician judgement on **any** domain (motor, cognitive, behavioural). Some manifest participants have DCL below 4.
- Codes 1 (unknown) and 6 (community control) never appear in a PDS.
:::
:::

## CAP: putting age and CAG on one scale

The CAG-Age Product summarises cumulative exposure to mutant huntingtin, like pack-years for smoking.

$$\text{CAP} = \text{age} \times \frac{\text{CAG} - 30}{6.49}$$

- Standardised so that **CAP = 100 at the expected age of motor diagnosis** (Warner et al. 2020). This is the `capscore` column in the PDS, computed at every visit for CAG 36 and above.
- Older variants exist: Zhang 2011 uses age × (CAG − 33.66) with stages early < 290, mid 290 to 367, late > 367. Say which one you use.
- CAP is undefined for people without the expansion. Do not compute it for controls.
- Enroll-HD quartile cut points for CAG 40+ (PDS4): 88 and 119, giving groups below 88, 88 to 119, above 119.

```{r}
#| label: cap-fn
cap_score <- function(age, cag) {
  # Warner et al. (2020) standardisation; NA for non-expanded alleles
  if_else(cag >= 36, age * (cag - 30) / 6.49, NA_real_)
}
cap_score(age = c(45, 45, 60), cag = c(42, 17, 40))
```

# 2. The Enroll-HD platform

## What Enroll-HD is

- A global, prospective, observational cohort study **and** a clinical research platform, run by CHDI Foundation since July 2012.
- All-comers: HD gene expansion carriers at any stage, plus genotype-negative relatives and family controls as comparators.
- Roughly 24,000 participants recruited, about 160 to 190 sites in over 20 countries across Europe, North America, Latin America and Australasia.
- **Annual in-person visits** with a mandatory core battery and optional extended assessments.
- Predecessor European study **REGISTRY** (2004 to 2016, protocol versions R2 and R3) migrated into the database for consenting participants, so many Europeans have history before their Enroll-HD baseline.

## What is measured at a visit

::: columns
::: {.column width="50%"}
**Core (every visit)**

- Demographics, HD clinical characteristics, medications, comorbidities, therapies
- UHDRS motor and diagnostic confidence
- UHDRS Total Functional Capacity, Functional Assessment, Independence Scale
- Cognitive: Symbol Digit Modalities, Stroop word / colour / interference, category fluency
- Problem Behaviours Assessment short form (PBA-s)
- Biosamples (CAG at baseline)
:::
::: {.column width="50%"}
**Extended and optional**

- Letter fluency, Trail Making A and B, MMSE
- HADS-SIS (anxiety, depression, irritability)
- Columbia Suicide Severity Rating Scale
- SF-12 health survey, WPAI work productivity
- Physiotherapy: timed up-and-go, chair stand
- Caregiver quality of life, service receipt inventory
- Reportable events: suicide attempt, suicide, mental-health hospitalisation, death
:::
:::

## Periodic datasets (PDS)

- Time-stamped cuts of the whole database, released every one to two years. PDS7 (2025) holds about **30,500 participants** and **130,000 visits**.
- Everyone in a PDS has a monitored baseline; most have follow-ups. Enrolment more than 15 months before the cut requires a further reviewed contact, so single-visit participants are mostly recent enrolments.
- Access: verify your institution, sign the data use agreement, post a short project description. Access is free and takes days, not months.
- **Specified datasets (SPS)** can be requested for variables suppressed from the PDS (site, dates, item-level questionnaires, family history), after scientific review.
- Public tier versus "available upon SRC approval" tier: the data dictionary tells you which.

## The documentation you should read first

| Document | Why it matters |
|---|---|
| Data dictionary (xlsx) | Every variable: file, form, type, code list, transformation, availability tier. The single source of truth |
| Explore Dataset Structure | Entity-relationship diagram, keys, visit ordering |
| Understand and Interpret Data | HD category rules, missing codes, date handling, score calculations, HD-ISS |
| Coding Systems | Drug, indication and comorbidity code formats |
| PDS overview | Sample sizes, distributions, completeness by form |
| Study protocol, annotated CRFs, data-collection guidelines | What was asked, in what order, with what skip logic |
| Unusual findings | Verified-correct oddities you will otherwise chase as bugs |

# 3. Anatomy of a PDS

## Eleven files, three grains

```{r}
#| label: files-table
#| echo: false
tribble(
  ~File, ~Grain, ~Key, ~Content,
  "profile",       "participant",          "subjid",                 "demographics, HD clinical characteristics, CAG, mortality",
  "pharmacotx",    "participant, repeating","subjid + row",          "medications with start/stop days, codes, ATC",
  "nutsuppl",      "participant, repeating","subjid + row",          "nutritional supplements",
  "nonpharmacotx", "participant, repeating","subjid + row",          "physiotherapy, speech therapy, counselling ...",
  "comorbid",      "participant, repeating","subjid + row",          "conditions (ICD-10) and procedures",
  "participation", "participant x study",  "subjid + studyid",       "status, category at entry and latest, visit list, end of study",
  "assessment",    "visit",                "subjid + studyid + seq", "1/blank flag per form completed at the visit",
  "event",         "reportable event",     "subjid + studyid + seq", "suicide attempt, suicide, hospitalisation, death",
  "enroll",        "Enroll-HD visit",      "subjid + studyid + seq", "all Enroll-HD visit forms and scores",
  "registry",      "REGISTRY visit",       "subjid + studyid + seq", "REGISTRY 2 and 3 visit forms",
  "adhoc",         "retrospective visit",  "subjid + studyid + seq", "UHDRS, cognitive and MMSE collected before REGISTRY"
) |>
  gt() |>
  tab_options(table.font.size = px(18)) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels())
```

## Studies and visits inside one PDS

::: columns
::: {.column width="50%"}
`studyid`

| code | study | when |
|---|---|---|
| ENR | Enroll-HD | mandatory, `visdy` from 0 |
| R3 | REGISTRY v3 | before ENR, negative `visdy` |
| R2 | REGISTRY v2 | before R3 |
| RET | Ad hoc / retrospective | before everything |
:::
::: {.column width="50%"}
`visit`

- Baseline, Follow Up, Unscheduled, Phone Contact, Retro Visit
- `seq` = 1 at baseline, then chronological, **including** unscheduled visits and phone contacts
- Phone contacts carry only the Missed Visit form; unscheduled visits carry the core battery but not MMSE, SF-12, HADS or WPAI
- `visdy` = days since the **Enroll-HD** baseline, for every study
:::
:::

## The keys in one picture

```{r}
#| label: erd
#| echo: false
#| fig-height: 4.5
nodes <- tribble(
  ~file, ~x, ~y, ~grain,
  "profile",        1.0, 3.0, "participant",
  "pharmacotx",     0.4, 2.0, "participant, repeating",
  "comorbid",       0.4, 1.5, "participant, repeating",
  "nutsuppl",       0.4, 1.0, "participant, repeating",
  "nonpharmacotx",  0.4, 0.5, "participant, repeating",
  "participation",  3.6, 3.0, "participant x study",
  "event",          3.6, 1.0, "event",
  "enroll",         6.0, 3.7, "visit",
  "registry",       6.0, 2.6, "visit",
  "adhoc",          6.0, 1.5, "visit",
  "assessment",     8.0, 2.6, "visit"
)
edges <- tribble(
  ~from, ~to, ~label,
  "profile", "participation", "subjid",
  "participation", "enroll", "subjid + studyid",
  "participation", "registry", "subjid + studyid",
  "participation", "adhoc", "subjid + studyid",
  "enroll", "assessment", "subjid + studyid + seq",
  "registry", "assessment", "",
  "adhoc", "assessment", ""
) |>
  left_join(nodes |> select(from = file, x0 = x, y0 = y), by = "from") |>
  left_join(nodes |> select(to = file, x1 = x, y1 = y), by = "to")

ggplot() +
  geom_segment(data = edges, aes(x = x0, y = y0, xend = x1, yend = y1),
               colour = "grey60") +
  geom_label(data = nodes, aes(x, y, label = file, fill = grain),
             size = 4.2, label.padding = unit(0.4, "lines")) +
  geom_text(data = edges |> filter(label != ""),
            aes((x0 + x1) / 2, (y0 + y1) / 2 + 0.18, label = label),
            size = 3, colour = "grey30") +
  scale_fill_brewer(palette = "Pastel1", name = NULL) +
  scale_x_continuous(expand = expansion(mult = 0.12)) +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(legend.position = "bottom")
```

# 4. Loading and merging

## Read everything as text first

The PDS ships as delimited text, one file per table, blank for missing. Aggregated cells such as `">70"` and `"<18"` sit inside otherwise numeric columns, so **read as character and convert deliberately**.

```{r}
#| label: read-pattern
#| eval: false
pds_dir <- here::here("data", "pds7")

read_pds <- function(name) {
  read_delim(
    file.path(pds_dir, paste0(name, ".csv")),
    delim = ",",                 # check: some releases are tab-delimited
    col_types = cols(.default = col_character()),
    na = c("", "NA"),
    show_col_types = FALSE
  )
}

pds <- c("profile", "participation", "enroll", "registry",
         "adhoc", "assessment", "event", "pharmacotx",
         "nutsuppl", "nonpharmacotx", "comorbid") |>
  set_names() |>
  map(read_pds)
```

For the rest of this deck `pds` is the toy list from `make_toy_pds()`, already typed.

## Participant tables and visit tables

```{r}
#| label: peek
pds$profile |> select(subjid, sex, caghigh, caglow, hddiagn)
pds$enroll |> select(subjid, seq, visit, visdy, age, hdcat, motscore, diagconf) |> head(6)
```

## Merging participant data onto visits

The join keys are `subjid` for participant tables and `subjid + studyid + seq` (or `subjid + visdy`) for visit tables. Say which grain the result has.

```{r}
#| label: merge
carriers <- pds$profile |>
  mutate(cag = parse_number(caghigh)) |>      # ">70" becomes 70; see next section
  filter(cag >= 36)

visits <- pds$enroll |>
  inner_join(carriers |> select(subjid, sex, cag), by = "subjid") |>
  left_join(pds$participation |> select(subjid, studyid, hdcat_0, age_0),
            by = c("subjid", "studyid")) |>
  mutate(years_in_study = visdy / 365.25)

visits |> select(subjid, seq, years_in_study, sex, cag, hdcat_0, motscore) |> head(5)
```

## The "last visit" and "first visit" patterns

```{r}
#| label: last-visit
last_visit <- pds$enroll |>
  group_by(subjid, studyid) |>
  slice_max(seq, n = 1, with_ties = FALSE) |>
  ungroup()

baseline <- pds$enroll |>
  filter(visit == "Baseline")          # equivalently seq == 1 and visdy == 0

bind_rows(baseline = baseline, latest = last_visit, .id = "which") |>
  count(which, hdcat)
```

Prefer `slice_max(seq)` over `filter(visdy == max(visdy))`: it is explicit about ties and reads as the intent.

## participation is wide; make it long

`participation` lists visits as `visit1 ... visit21` and `vis1dy ... vis21dy`. Reshape before you count.

```{r}
#| label: participation-long
#| eval: false
visit_list <- pds$participation |>
  select(subjid, studyid, starts_with("visit"), matches("^vis\\d+dy$")) |>
  pivot_longer(
    cols = -c(subjid, studyid),
    names_to = c(".value", "index"),
    names_pattern = "(visit|vis)(\\d+)(?:dy)?"
  ) |>
  rename(visit_type = visit, visit_day = vis) |>
  filter(!is.na(visit_type)) |>
  mutate(index = as.integer(index), visit_day = as.integer(visit_day))
```

Visit types there are abbreviated: BL, FUP, PC (phone contact), U (unscheduled), R (ad hoc), E (premature end).

# 5. De-identification conventions

## No dates, only days and ages

- Every date became a **day offset from the Enroll-HD baseline**: `visdy`, `cmstdy`, `cmendy`, `mhstdy`, `evtdy`, `rfendy`, ... Negative means before baseline.
- Every date of birth became an **integer age**: `age_0`, `age` at each visit, `hddiagn`, `dssage`, parental onset ages.
- Partial dates were completed before conversion: a missing day became the 15th, a missing month and day became 1 July. The completion flag is not released.

::: callout-warning
## Consequence
Start and stop days of a medication or condition can be **equal or reversed**. A duration of zero or minus 14 days is an artefact, not an error. Do not "clean" it away; model durations with that in mind.
:::

## Aggregation: numbers that arrive as text

| Column | Aggregated values |
|---|---|
| `age`, `age_0` | `"<18"`, `">90"` |
| `caghigh` | `">70"` |
| `caglow` | `">28"` |
| `bmi_imp` | `"<16.7"`, `">43.7"` |
| `sbh1n` (suicide attempts) | `">50"` |
| `race` | collapsed to seven categories |

```{r}
#| label: parse-aggregated
parse_aggregated <- function(x) {
  # Keep the numeric part, and flag cells that were censored by aggregation
  tibble(value = parse_number(x),
         censored = case_when(str_starts(x, "<") ~ "below",
                              str_starts(x, ">") ~ "above",
                              TRUE ~ "exact"))
}

pds$profile |>
  mutate(parse_aggregated(caghigh)) |>
  select(subjid, caghigh, value, censored) |>
  filter(censored != "exact")
```

## Two kinds of missing

::: columns
::: {.column width="50%"}
**System-defined**: a blank cell (`NA`). A question that was skipped by design, a child of a "no" answer, or a total with a missing item.

**User-defined**: a code inside the column.

| numeric | text | meaning |
|---|---|---|
| 9996 | WRONG | value known to be wrong |
| 9997 | NOTAPPL | not applicable |
| 9998 | MISSING | refused or omitted |
| 9999 | UNKNOWN | unknown to participant |
:::
::: {.column width="50%"}
```{r}
#| label: missing-codes
missing_codes <- c(9996, 9997, 9998, 9999)

pds$enroll |>
  mutate(across(c(motscore, tfcscore, sdmt1),
                ~ if_else(.x %in% missing_codes, NA, .x))) |>
  summarise(across(c(motscore, tfcscore, sdmt1),
                   ~ sum(is.na(.x))))
```

Do this **before** any arithmetic; 9998 in a mean is a silent disaster.
:::
:::

## Other things you will notice

- `subjid` is a one-way recoded ID, `R` plus nine digits. Stable within a release, **not** across releases.
- Height and weight are withheld; `bmi_imp` uses baseline height at every visit and is blank under 18.
- Site, country and exact dates are not in the public tier.
- Genotype-unknown participants were reclassified using the research CAG; community controls were removed.
- Comorbidities, therapies and events had `studyid` removed because they are study-independent.
- Exact duplicate rows in medication and comorbidity files are **intentional** (re-recorded entries). Decide explicitly whether to `distinct()`.

# 6. Disease definitions and derived scores

## Scores you can recompute

| Score | Definition | Range |
|---|---|---|
| `motscore` | sum of 31 UHDRS motor items; blank if any item missing, then `miscore` holds the partial sum | 0 to 124 |
| `tfcscore` | occupation + finances + chores + ADL + care level | 0 to 13 |
| `fascore` | count of 25 "yes" items; `fiscore` if incomplete | 0 to 25 |
| `indepscl` | single item | 10 to 100, step 5 |
| PBA-s domains | severity × frequency summed over items (depression 1 to 3, irritability 4 to 5, psychosis 9 to 10, apathy 6, executive 7 to 8) | |
| `capscore` | age × (CAG − 30) / 6.49 | |

```{r}
#| label: shoulson
shoulson_fahn_stage <- function(tfc) {
  case_when(tfc >= 11 ~ 1L, tfc >= 7 ~ 2L, tfc >= 3 ~ 3L,
            tfc >= 1 ~ 4L, tfc == 0 ~ 5L)
}
pds$enroll |> mutate(stage = shoulson_fahn_stage(tfcscore)) |> count(stage)
```

## Deriving motor onset within the study

```{r}
#| label: motor-onset
motor_onset <- pds$enroll |>
  group_by(subjid) |>
  arrange(seq, .by_group = TRUE) |>
  summarise(
    dcl4_at_entry     = first(diagconf) == 4,
    converted         = !dcl4_at_entry & any(diagconf == 4),
    onset_visdy       = if_else(converted, visdy[which(diagconf == 4)[1]], NA_integer_),
    last_visdy        = last(visdy),
    .groups = "drop"
  )
motor_onset
```

- Participants already at DCL 4 at baseline have **no observable onset**; exclude them from time-to-onset analyses or use methods that handle left truncation.
- The same pattern with `hdcat` (2 to 3) gives clinical onset in any domain.
- `hddiagn` is the age the participant was **told**; it can lag onset by years or be missing even for manifest participants.

## Check the category rules in your data

Rules from the documentation, written as tests. Run them once on every new release.

```{r}
#| label: rule-checks
category_checks <- pds$participation |>
  left_join(pds$profile |> mutate(cag = parse_number(caghigh)) |>
              select(subjid, cag), by = "subjid") |>
  summarise(
    controls_below_36   = all(cag <  36 | !hdcat_0 %in% c(4, 5)),
    carriers_36_or_more = all(cag >= 36 | !hdcat_0 %in% c(2, 3)),
    no_unknown_or_cc    = !any(hdcat_0 %in% c(1, 6)),
    latest_not_earlier  = all(hdcat_l >= hdcat_0 | !hdcat_0 %in% c(2, 3))
  )
category_checks
```

# 7. Special characteristics of HD data

## Floors and ceilings everywhere

```{r}
#| label: toy-cohort
#| echo: false
# A larger illustrative cohort for the distribution and modelling slides.
# Shapes follow the documentation qualitatively; values are invented.
set.seed(7)
toy_cohort <- tibble(subjid = sprintf("R%09d", 1:400)) |>
  mutate(
    cag  = sample(38:50, n(), replace = TRUE,
                  prob = c(1, 2, 6, 10, 12, 11, 9, 6, 4, 3, 2, 1, 1)),
    age  = round(runif(n(), 22, 70)),
    cap  = cap_score(age, cag),
    manifest = rbinom(n(), 1, plogis((cap - 100) / 12)) == 1,
    motscore = if_else(manifest,
                       rnbinom(n(), mu = pmax(5, (cap - 70) * 0.9), size = 3),
                       rnbinom(n(), mu = 1.5, size = 0.6)) |>
      pmin(124),                                  # the UHDRS motor maximum
    tfcscore = if_else(manifest,
                       pmin(13, pmax(0, round(13 - (cap - 90) / 8 + rnorm(n(), 0, 1.5)))),
                       pmin(13, round(13 - rexp(n(), 3)))),
    education_years = round(rnorm(n(), 13, 3))
  )
```

```{r}
#| label: floor-ceiling
#| echo: false
#| fig-height: 3.8
toy_cohort |>
  select(subjid, manifest, motscore, tfcscore) |>
  pivot_longer(c(motscore, tfcscore), names_to = "score") |>
  mutate(group = if_else(manifest, "manifest", "premanifest")) |>
  ggplot(aes(value, fill = group)) +
  geom_histogram(bins = 30, position = "identity", alpha = 0.6) +
  facet_wrap(~ score, scales = "free") +
  labs(x = NULL, y = "visits", fill = NULL,
       caption = "Illustrative data. Real distributions are in the PDS overview and enroll-hd.org.") +
  theme(legend.position = "bottom")
```

- Total motor score piles up at **0** (about a quarter of baseline HDGEC visits in the lowest CAP quartile) and is right-skewed to 124.
- TFC, FAS and the Independence Scale pile up at their **maximum** for early participants.
- Consequence: means and t-tests mislead. Consider transformations, zero-inflated or negative-binomial models for TMS, ordinal or beta-type models for TFC, or model change rather than level.

## Behaviour behaves differently

- PBA-s domains (depression, irritability, psychosis, apathy, executive) track disease stage **weakly**; apathy is the exception and rises with progression.
- Behavioural items are informant-dependent: the "worst" ratings are blank at the Enroll-HD baseline, and several items are 9 (not assessable) when the participant came alone.
- Suicidal ideation on PBA-s should agree with the C-SSRS rows and the event file. Check it.

## Selection and follow-up

- Cohort skews toward higher socioeconomic status and better health; ethnic diversity is limited (about 92% Caucasian).
- Retention: about **80%** of premanifest and **55%** of manifest carriers remain at 7 years (withdrawal, death, loss to follow-up).
- Visits are annual in intent: median gap **374 days**, but skewed, with gaps up to 6.8 years. People skip years and return, or pause during clinical trials.

```{r}
#| label: visit-gaps
pds$enroll |>
  group_by(subjid) |>
  arrange(seq, .by_group = TRUE) |>
  mutate(gap_days = visdy - lag(visdy)) |>
  ungroup() |>
  summarise(n_gaps = sum(!is.na(gap_days)),
            median_gap = median(gap_days, na.rm = TRUE),
            longest_gap = max(gap_days, na.rm = TRUE))
```

# 8. Age, CAG and time in models

## Cross-sectional: index progression with CAP

At study entry, disease burden differs by age **and** CAG. Enter both, or their product.

```{r}
#| label: cross-sectional
#| eval: false
baseline_carriers <- pds$enroll |>
  filter(visit == "Baseline") |>
  inner_join(pds$profile |> mutate(cag = parse_number(caghigh)) |>
               filter(cag >= 40), by = "subjid") |>
  mutate(age = parse_number(age), cap = cap_score(age, cag))

fit_cap <- lm(sdmt1 ~ cap + sex + education_years, data = baseline_carriers)
broom::tidy(fit_cap, conf.int = TRUE)
```

- Education matters for cognitive outcomes; sex interactions matter in some contexts.
- CAP groups (below 88, 88 to 119, above 119) are a defensible early/mid/late split for CAG 40+.
- Beware confounding by CAG: in Enroll-HD, participants with a history of drug use have milder motor signs **only because they have shorter CAG expansions**.

## Longitudinal: use time since entry, not age, as the clock

```{r}
#| label: lmm
#| eval: false
library(lme4)

carrier_visits <- visits |>
  filter(cag >= 40) |>
  mutate(cap_group = cut(cap_score(parse_number(age_0), cag),
                         breaks = c(-Inf, 88, 119, Inf),
                         labels = c("early", "mid", "late")))

fit_slopes <- lmer(
  motscore ~ years_in_study * cap_group + sex + (1 + years_in_study | subjid),
  data = carrier_visits
)
broom.mixed::tidy(fit_slopes, effects = "fixed", conf.int = TRUE)
```

- Time 0 = the baseline visit; CAP (or CAP group) carries progression **up to** entry. Within a CAP group, change over a few years is close to linear.
- Over the whole adult lifespan trajectories are not linear; if age is the clock use splines or fractional polynomials.
- Random slopes are the norm: people progress at different rates.

## Time to a landmark: survival methods

```{r}
#| label: survival
#| eval: false
library(survival)

onset_data <- motor_onset |>
  filter(!dcl4_at_entry) |>
  inner_join(pds$profile |> mutate(cag = parse_number(caghigh)), by = "subjid") |>
  inner_join(pds$participation |> select(subjid, age_0), by = "subjid") |>
  mutate(
    time_years = coalesce(onset_visdy, last_visdy) / 365.25,
    event      = as.integer(converted),
    cap_entry  = cap_score(parse_number(age_0), cag)
  )

fit_cox <- coxph(Surv(time_years, event) ~ cap_entry + sex, data = onset_data)
broom::tidy(fit_cox, exponentiate = TRUE, conf.int = TRUE)
```

- Exclude people who were already past the landmark at entry, and think about whether that exclusion biases you.
- Joint longitudinal-survival models exist for using all visits; see Long and Mills (2018).

## Signal, noise and clinical trials

Enroll-HD is used to design trials. The quantity that matters is the outcome's **signal-to-noise ratio**: mean annual change divided by the within-person SD of annual change.

- Example from the articles: TFC declines 1.2 points/year with within-person SD 0.8, so signal-to-noise is 1.5. A treatment halving the decline has effect size 0.75.
- Sample size scales with the **square** of that ratio, and with the square of follow-up length.
- Estimate signal and noise on the subset of participants that a trial would enrol, and on the first few years of follow-up only.
- Baseline prognostic covariates (CAP, baseline score) reduce noise; **enrichment** selects those likely to progress during the trial.

# 9. Coded therapies and comorbidities

## Codes in the current PDS

| What | Column | Format |
|---|---|---|
| Drug | `cmtrt__decod` | RX + 9 digits (internal); `cmtrt__modify` is the English term |
| Active ingredients | `cmtrt__ing` | comma-separated |
| ATC classes | `cmtrt__atc` | comma-separated ATC codes |
| Indication | `cmindc__decod` | CX + 9 digits; `cmindc__modify` English |
| Comorbidity | `mhterm__decod` | ICD-10 (2014) |
| Procedure | `mhterm__decod` | CX + 9 digits |
| Ongoing | `cmenrf`, `mhenrf` | 1 if and only if the stop day is blank |

Older releases used WHO Drug Dictionary and MedDRA codes; the format changed, the logic did not.

## Filter by therapeutic class, not by name

```{r}
#| label: atc-filter
on_antidepressant <- pds$pharmacotx |>
  mutate(atc_codes = str_split(cmtrt__atc, ",\\s*")) |>
  filter(map_lgl(atc_codes, ~ any(str_starts(.x, "N06A")))) |>
  mutate(ongoing = cmenrf == 1,
         duration_days = cmendy - cmstdy) |>
  select(subjid, cmtrt__modify, cmtrt__atc, cmstdy, cmendy, ongoing, duration_days)
on_antidepressant
```

Note `R000000002`: an exact duplicate row and a zero-day duration from partial-date imputation. Both are real-data features.

## Comorbidities and ICD-10 chapters

```{r}
#| label: icd-filter
#| eval: false
depression_history <- pds$comorbid |>
  filter(str_starts(mhterm__decod, "F32|F33")) |>
  distinct(subjid) |>
  mutate(depression_ever = TRUE)

pds$profile |>
  left_join(depression_history, by = "subjid") |>
  mutate(depression_ever = coalesce(depression_ever, FALSE)) |>
  count(depression_ever)
```

- ICD-10 is hierarchical: `str_starts` on the chapter or block prefix gives you the family of codes.
- `mhbodsys` gives a coarse body-system code (1 to 17) when you just need organ systems.
- Comorbidity and therapy rows are participant-level, not visit-level: align them to visits with `cmstdy`/`cmendy` and `visdy`.

# 10. Data quality and a pre-analysis checklist

## Quality you get, and quality you must add

**Provided**: EDC edit checks and auto-computed totals, remote and on-site monitoring, central statistical monitoring, central coding, a published list of unusual but verified findings.

**Yours to do**:

- Recode 9996 to 9999 to `NA` before any arithmetic.
- Parse aggregated text columns explicitly and keep the censoring flag.
- Recompute totals from items where items are available and compare.
- Confirm one baseline per participant at `visdy` 0 and `seq` 1, increasing `visdy`, and that `visitnum` equals the number of visit rows.
- Look for known oddities before calling them bugs: pack-years of 2005, parental onset age 0, zero-length prescriptions.
- Decide and document how to treat duplicates, unscheduled visits and phone contacts.

## A tidy project for Enroll-HD analyses

```
my-enrollhd-analysis/
  my-enrollhd-analysis.Rproj
  renv.lock                 # frozen package versions
  data/                     # the PDS extract, git-ignored, never copied around
  R/
    read_pds.R              # one function per concern
    recode_missing.R
    derive_scores.R
    build_analysis_set.R
  analysis/
    01_cohort.qmd           # one Quarto document per analysis question
    02_progression.qmd
  output/                   # figures and tables, regenerated, git-ignored
  SAP.md                    # the statistical analysis plan, written first
```

- Functions take a tibble and return a tibble. Pipelines read top to bottom. Names say what, comments say why.
- The analysis set is built by code from the raw PDS every time; no hand edits.

## Before you begin (from the Enroll-HD statisticians)

::: {.incremental}
1. Write the research question and objectives down, then search the literature.
2. Define the population of interest and how it is operationalised in PDS variables.
3. Check feasibility: are there enough of those people, with enough visits?
4. Write a statistical analysis plan: dataset and release, measures, cleaning and outlier rules, missing-data strategy, methods, covariates and modifiers (CAG × age, sex, education), multiplicity, reproducibility. Consider pre-registering it.
5. Work with a statistician throughout, not only at the end.
6. Cite the release version and the documentation you relied on.
:::

## Where to go next

- Analyzing Data hub: <https://www.enroll-hd.org/for-researchers/analyzing-data/>
  - Special characteristics of HD data; Age and CAG length in HD data analysis; Data handling and management tips; Benefits and challenges of observational data; Using observational data to inform clinical trial design; Enrichment in clinical trial recruitment; Before you begin.
- Documentation hub (dictionary, protocol, CRFs, coding, overview): <https://www.enroll-hd.org/for-researchers/data-support-documentation/>
- Tools: PDS Data Explorer, HD-ISS calculator, Atlas of HD phenotype (same site).
- Key papers: Landwehrmeyer et al. 2017 (Enroll-HD design); Warner et al. 2020 (CAP); Long & Mills 2018 (joint models); Schobel et al. 2017 (cUHDRS); Langbehn et al. 2010 (CAG and onset); Tabrizi et al. 2013 (TRACK-HD).

Questions, corrections and additions are welcome. Source, toy data and rendering instructions: <https://github.com/stat-absk/enrollhd-getting-started>.
